Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression

### 2022-06-26 docs/rmd/presentation/slides.Rmd
autogit
kinda done

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
---

```{r setup, include=FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

knitr::kable(
  read.csv("../tables/csv/gp_parameters.csv"),
  digits=3,
  caption="Evolutionary Parameters"
)
```

## Function Set

@@

knitr::kable(
  read.csv("../tables/csv/primitives.csv"),
  digits=3,
) 
```

## Terminal Set

@@

knitr::kable(
  read.csv("../tables/csv/terminals.csv"),
  digits=3,
) 
```

@@

### 3 Questions

1. Differences in average fitness for both algorithms?
2. Differences in fitness for training and testing data?
3. Differences in program size?

@@

@@

* The gap between training and testing error is not statistically significant for both selection algorithms
* Both algorithms achieve a slightly better performance for the training data
* No proof for overfitting

@@

```{r, echo=FALSE, warning=FALSE}

df <- read.csv("../tables/csv/mwu_matrix_error.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Mean Error - P-Values (MWU)"
)

@@

@@

...

@@

...

@@

...
\newpage

>>>>

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
header-includes:
   - \usepackage{float}
---

```{r setup, include=FALSE}
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

knitr::kable(
  read.csv("../tables/csv/gp_parameters.csv"),
  format="latex",
  digits=3,
  caption="Evolutionary Parameters"
)
```

## Primitive Set {.allowframebreaks}

@@

knitr::kable(
  read.csv("../tables/csv/primitives.csv"),
  format="latex",
  digits=3,
  caption="Function Set"
) 
```
\framebreak

@@

knitr::kable(
  read.csv("../tables/csv/terminals.csv"),
  format="latex",
  digits=3,
  caption="Terminal Set"
) 
```

@@

### Statistical Testing Strategy

1. Differences in average fitness for both algorithms on both datasets?
2. Differences in fitness for training and testing data?

@@

* The gap between training and testing error is not statistically significant for both selection algorithms
* Both algorithms achieve a slightly better performance for the training data
* Good generalization: No evidence of overfitting

@@

@@

```{r, echo=FALSE, warning=FALSE}
df <- read.csv("../tables/csv/mwu_matrix_error.csv")

knitr::kable(
  df,
  format="latex",
  digits=3,
  caption="Mean Error - P-Values (MWU)"
) %>% kableExtra::kable_styling(font_size = 4)

@@

## Program Growth

* So far: No proof of differences in generalization
* New approach: Program growth as a possible indicator for overfitting?
* Theory: Minimum description length principle (MDLP) ^[@wang_wagner_rondinelli_2019, p. 268]
* Downside: Growth/Bloat is no clear indicator of overfitting ^[@bloat_overfitting_gp, p. 8]

@@

* GP typical growth behaviour for both operators
* Solutions grow at a similiar rate in each generation
* No statistically significant differences in overall program size based on selection

@@

* Experiment did not yield evidence for differences in generalization behavior between tournament and $\epsilon$-lexicase selection
* The performance of tournament selection is significantly higher than that of $\epsilon$-lexicase selection for the selected symbolic regression problem
* No evidence for differences in growth behavior between both algorithms

@@

1. Configuration of evolutionary parameters
2. Results are based on a single symbolic regression
2. Limited by computational resources

# References {.allowframebreaks}

### 2022-06-26 docs/rmd/presentation/slides.tex
kinda done

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ignorenonframetext,
]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{part title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{part title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usecolortheme{beaver}
\usefonttheme{structurebold}
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\newif\ifbibliography
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
\centering
\begin{tabular}[t]{l|l}
\hline
Variable & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Function Set}
\protect\hypertarget{function-set}{}
\begin{tabular}{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{frame}

\begin{frame}{Terminal Set}
\protect\hypertarget{terminal-set}{}
\begin{tabular}{l|l}
\hline
Terminal & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
random\_int & Ephemeral Constant (integer)\\
\hline
random\_float & Ephemeral Constant(float)\\
\hline
\end{tabular}
\end{frame}

@@

\begin{block}{3 Questions}
\protect\hypertarget{questions}{}
\begin{enumerate}
\tightlist
\item
  Differences in average fitness for both algorithms?
\item
  Differences in fitness for training and testing data?
\item
  Differences in program size?
\end{enumerate}
\end{block}
\end{frame}

@@

@@

\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\midrule
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Finding 2}
\protect\hypertarget{finding-2}{}
\ldots{}
\end{frame}

\begin{frame}{Evolution of Size}
\protect\hypertarget{evolution-of-size}{}
\begin{figure}
\centering
\includegraphics{../plots/size_subplotted.png}
\end{figure}

\ldots{}
\end{frame}

@@

\begin{frame}{Conclusions}
\ldots{}
\end{frame}

@@

\begin{frame}{Limitations and open Questions}
\ldots{} \newpage

\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Dua:2019}{}%
Dua, D. and Graff, C. (2017) `UCI machine learning repository'.
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

@@

\leavevmode\hypertarget{ref-Orzechowski_2018}{}%
Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) `Where are we now?',
in \emph{Proceedings of the genetic and evolutionary computation
conference}. ACM.
doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.
\end{cslreferences}
\end{frame}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  ignorenonframetext,
]{beamer}
\usepackage{pgfpages}
\setbeamertemplate{caption}[numbered]
\setbeamertemplate{caption label separator}{: }
\setbeamercolor{caption name}{fg=normal text.fg}
\beamertemplatenavigationsymbolsempty
% Prevent slide breaks in the middle of a paragraph
\widowpenalties 1 10000
\raggedbottom
\setbeamertemplate{part page}{
  \centering
  \begin{beamercolorbox}[sep=16pt,center]{part title}
    \usebeamerfont{part title}\insertpart\par
  \end{beamercolorbox}
}
\setbeamertemplate{section page}{
  \centering
  \begin{beamercolorbox}[sep=12pt,center]{part title}
    \usebeamerfont{section title}\insertsection\par
  \end{beamercolorbox}
}
\setbeamertemplate{subsection page}{
  \centering
  \begin{beamercolorbox}[sep=8pt,center]{part title}
    \usebeamerfont{subsection title}\insertsubsection\par
  \end{beamercolorbox}
}
\AtBeginPart{
  \frame{\partpage}
}
\AtBeginSection{
  \ifbibliography
  \else
    \frame{\sectionpage}
  \fi
}
\AtBeginSubsection{
  \frame{\subsectionpage}
}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usecolortheme{beaver}
\usefonttheme{structurebold}
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression},
  pdfauthor={Roman Hoehn, B.Sc. Wirtschaftspaedagogik},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\newif\ifbibliography
\usepackage{longtable,booktabs}
\usepackage{caption}
% Make caption package work with longtable
\makeatletter
\def\fnum@table{\tablename~\thetable}
\makeatother
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {}%
  {\par}

@@

\begin{longtable}[]{@{}ll@{}}
\caption{Overview - Energy Heating data set}\tabularnewline
\toprule
Variable & Description\tabularnewline
\midrule
\endfirsthead
\toprule
Variable & Description\tabularnewline
\midrule
\endhead
X1 & Relative Compactness\tabularnewline
X2 & Surface Area\tabularnewline
X3 & Wall Area\tabularnewline
X4 & Roof Area\tabularnewline
X5 & Overall Height\tabularnewline
X6 & Orientation\tabularnewline
X7 & Glazing Area\tabularnewline
X8 & Glazing Area Distribution\tabularnewline
y1 & Heating Load\tabularnewline
y2 & Cooling Load\tabularnewline
\bottomrule
\end{longtable}
\end{frame}

@@

\begin{frame}[allowframebreaks]{Primitive Set}
\protect\hypertarget{primitive-set}{}
\begin{table}

\caption{\label{tab:unnamed-chunk-3}Functions}
\centering
\begin{tabular}[t]{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{table}
\framebreak

\caption{\label{tab:unnamed-chunk-4}Terminals}
\centering
\begin{tabular}[t]{l|l}
\hline
Terminal & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
random\_int & Ephemeral Constant (integer)\\
\hline
random\_float & Ephemeral Constant(float)\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{block}{Statistical Testing Strategy}
\protect\hypertarget{statistical-testing-strategy}{}
\begin{enumerate}
\tightlist
\item
  Differences in average fitness for both algorithms on both datasets?
\item
  Differences in fitness for training and testing data?
\end{enumerate}
\end{block}
\end{frame}

@@

\begin{frame}{Finding 2}
\protect\hypertarget{finding-2}{}
\begin{itemize}
\tightlist
\item
  The gap between training and testing error is not statistically
  significant for both selection algorithms
\item
  Both algorithms achieve a slightly better performance for the training
  data
\item
  Good generalization: No evidence of overfitting
\end{itemize}
\end{frame}

@@

\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
\centering
\fontsize{4}{6}\selectfont
\begin{tabular}[t]{l|r|r|r|r}
\hline
X & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\hline
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
\hline
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
\hline
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
\hline
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Program Growth}
\protect\hypertarget{program-growth}{}
\begin{itemize}
\tightlist
\item
  So far: No proof of differences in generalization
\item
  New approach: Program growth as a possible indicator for overfitting?
\item
  Theory: Minimum description length principle (MDLP)\footnote<.->{Wang,
    Wagner and Rondinelli (2019), p.~268}
\item
  Downside: Growth/Bloat is no clear indicator of
  overfitting\footnote<.->{Silva and Vanneschi (2009), p.~8}
\end{itemize}
\end{frame}

\begin{frame}{Evolution of Size}
\protect\hypertarget{evolution-of-size}{}
\begin{figure}
\centering
\includegraphics{../plots/size_subplotted.png}
\end{figure}
\end{frame}

\begin{frame}{Finding 3}
\protect\hypertarget{finding-3}{}
\begin{itemize}
\tightlist
\item
  GP typical growth behaviour for both operators
\item
  Solutions grow at a similiar rate in each generation
\item
  No statistically significant differences in overall program size based
  on selection
\end{itemize}
\end{frame}

@@

\begin{frame}{Conclusions}
\begin{itemize}
\tightlist
\item
  Experiment did not yield evidence for differences in generalization
  behavior between tournament and \(\epsilon\)-lexicase selection
\item
  The performance of tournament selection is significantly higher than
  that of \(\epsilon\)-lexicase selection for the selected symbolic
  regression problem
\item
  No evidence for differences in growth behavior between both algorithms
\end{itemize}
\end{frame}

@@

\begin{frame}{Limitations and open Questions}
\begin{enumerate}
\tightlist
\item
  Configuration of evolutionary parameters
\item
  Results are based on a single symbolic regression
\item
  Limited by computational resources
\end{enumerate}
\end{frame}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\begin{frame}[allowframebreaks]{References}
\hypertarget{refs}{}
\begin{cslreferences}
\leavevmode\hypertarget{ref-Dua:2019}{}%
Dua, D. and Graff, C. (2017) `UCI machine learning repository'.
University of California, Irvine, School of Information; Computer
Sciences. Available at: \url{http://archive.ics.uci.edu/ml}.

@@

\leavevmode\hypertarget{ref-Orzechowski_2018}{}%
Orzechowski, P., Cava, W.L. and Moore, J.H. (2018) `Where are we now?',
in \emph{Proceedings of the genetic and evolutionary computation
conference}. ACM.
doi:\href{https://doi.org/10.1145/3205455.3205539}{10.1145/3205455.3205539}.

\leavevmode\hypertarget{ref-bloat_overfitting_gp}{}%
Silva, S. and Vanneschi, L. (2009) `Operator equalisation, bloat and
overfitting: A study on human oral bioavailability prediction', in
\emph{Proceedings of the 11th Annual Genetic and Evolutionary
Computation Conference, GECCO-2009}, pp. 1115--1122.
doi:\href{https://doi.org/10.1145/1569901.1570051}{10.1145/1569901.1570051}.

\leavevmode\hypertarget{ref-wang_wagner_rondinelli_2019}{}%
Wang, Y., Wagner, N. and Rondinelli, J.M. (2019) `Symbolic regression in
materials science', \emph{MRS Communications}, 9(3), pp. 793--805.
doi:\href{https://doi.org/10.1557/mrc.2019.85}{10.1557/mrc.2019.85}.
\end{cslreferences}
\end{frame}
